Impacts of the AUMF Resolutions on Obama and Trump Administration Foreign Policy

### 2020-12-07 paper.tex
obama and trump admin first draft

\section*{Origins of the Bush Administration AUMF Resolutions}
Following the terrorist attacks against the United States that took place on September 11th, 2001, President George W. Bush role as the acting Commander in Chief and spokesman for the United States was tasked with both identifying the perpetrators of the attacks as well as deciding what counter-measures to take as a result.
President Bush identified two main groups as the parties responsible for the attacks: Al Qaeda and the Taliban.
It was believed that Al Qaeda was the group that planned and initiated the attacks, while the Afghani Taliban were also guilty due to harboring Al Qaeda members within their territory.
With the purposes of providing an immediate response that would effectively utilize Congress's constitutional war powers, the 107th Congress passed an Authorization for Use of Military Force (AUMF) resolution, similar to the one passed in 1991 under the George H.W. Bush administration.
This resolution was passed through both the Senate and House in identical form on September 14th, just three days after the attacks \autocite{zotero-165}.
Although that length of time is notably fast compared to many other landmark pieces of legislation passed through Congress, there was still a lot of discussion between the legislators both of the political parties, leaders in the Bush administration, as well as legal and foreign policy experts.

@@

By doing this, President Bush framed the attacks as more than just a tragic occurrence, they were a direct declaration of war on the free world.
In response to Al Qaeda's act of war, the Bush administration decided to push for a legal authorization for a war against the 9/11 perpetrators.
Vice President Dick Cheney consulted with his lawyers along with members of the White House Counsel and the Department of Justice Legal Counsel to see what kind of authority would be needed.
Following this, President Bush met with congressional leaders to specifically request an resolution giving an authorization for military force to the executive branch.
Additionally, Bush also had Vice President Cheney's team of counsel members create a first draft of the resolution \autocite[176]{murray2015}.
This draft was considered by Senate Majority Leader, who would later be the sponsor for the 2001 AUMF resolution, as a blank check that would allow the Bush administration to use any amount of military force without consequence.
Congress decided to pass an AUMF under the conditions that it would limited to those responsible for the attacks, as well as that it would remain consistent with the 1973 War Powers Resolution.
The Bush Administration tried to add a clause to the draft resolution that would allow for military action \emph{within} the United States, but it was rejected by Congress.
After the members of Congress agreed to a set of conditions along with the wording of the resolution, they passed the resolution through both the House and Senate in approximately 28 minutes with only a single vote in dissent \autocite[177]{murray2015}.

Although Congress went through lengthy discussions in order to make sure the wording of the resolution wouldn't have unforeseen consequences, there were still some parts that ending up being broader than intended.
The 2001 AUMF authorized the President to have access to Necessary and Proper clause in Article 1 of the Constitution in order to both combat and detain the perpetrators of the 9/11 attacks, and as such gave President Bush a large amount of military power that would typically be in the hands of Congress.
The first example is the phrase ``organizations, or persons'' which can be found in the first operative clause of the resolution.
This was dissimilar from Congress's historical declarations of war as well as the 1991 Authorization for Use of Military Force Against Iraq Resolution, which not only specifically chose the government of Iraq as a target, but also was much more clear in its goals.
Additionally, the 2001 AUMF resolution was somewhat broad in the way it described which non-state actors and/or nations should be targeted.
Malcolm Brooks Savage III describes how there was only one limitation placed on the powers of the President in the operational clauses of the 2001 AUMF resolution, being the 9/11 nexus requirement.
The first operative clause states that the President can use force against all ``nations, organizations, or persons \emph{he determines} planned, authorized, committed, or aided the terrorist attacks.''
The phrase ``he determines'' is especially notable as it could presumably allow the President to target any group that was remotely connected to the attacks, even if they are outside of the original Congressional intent of the resolution \autocite[5]{savage2015}.

The shift of military focus from state to non-state actors marked a drastic change within the foreign policy of the United States.
Although the Bush Administration specifically identified Al Qaeda and the Taliban as the targets of their response plans, the language of the 2001 AUMF resolution gave additional power to the President that went beyond those two groups specifically.
Additionally, the phrasing led to interpretations that would allow military force to be exerted not just on group actors, but also individuals.
As a result of this, President Bush as well as all future presidents that chose to implement the 2001 AUMF into their foreign policy decisions were able to choose targets in both a wider scope due to the lack of specificity and increased executive discretion, as well as choose targets that were much more specific, such as specific people that were deemed as terrorists \autocite[3]{weed2015}.
As will be discussed later in this paper, these parts of the 2001 AUMF, regardless of whether they should be considered features or loopholes, have had many direct consequences on the foreign policy of the United States for the last two decades.

@@

\begin{displayquote}
I began to have some qualms over how broad a grant of authority Congress
gave him in our rush to act quickly.
Because of the speed with which it was passed, there was little discussion establishing a foundation for the resolution.
Because of the paucity of debate, it would be difficult to glean from the
record the specific intent of Congress in approving S.J. Res. 23.
There were after-the-fact statements made in the Senate, and there was some debate in the House, but there was not the normal level of discussion or the normal
level of analysis of the language prior to the vote that we have come to expect in the Senate \autocite[S9949]{byrd2001a}.
\end{displayquote}

This proves how despite the level of deliberation taken, it was still rushed far beyond the amount expected out of such an important piece of legislation.
Although there was supposedly some benefit to this, being that the situation at hand demanded immediate action, this gives evidence to the idea that the text of the resolution wasn't completely aligned with the interests or intents of Congress.

\begin{displayquote}
Two aspects of the resolution are key: First, the use of force authority granted to the President extends only to the perpetrators of the September 11 attack.
It was not the intent of Congress to give the President unbridled authority—I hope it wasn’t to wage war against terrorism writ large without the advice and consent of Congress.
That intent was made clear when Senators modified the text of the resolution proposed by the White House to limit the grant of authority to the September 11 attack \autocite[S9949]{byrd2001a}.
\end{displayquote}

This quote describes the intent behind the 9/11 nexus requirement that was discussed by Savage earlier.
Later in the speech, Senator Byrd elaborates on the resolution draft that was proposed by the White House.
This draft, in addition to all of the powers granted in the version of the resolution which was later passed by Congress, would have allowed for the President to ``deter and pre-empt any future acts of terrorism or aggression against the United States \autocite[S9949]{byrd2001a}.''
The rejection of this proposed clause proves the congressional intent to limit the scope of the authorization.

Beyond this, the resolution also aimed to further the restrictions on the power of the executive branch by invoking the War Powers Resolution of 1973.
Section (2) (b) (1) of the resolution requires that specific statutory authorization should follow the meaning of section (5) (b) of the War Powers Resolution \autocite{daschle2001}.
The War Powers Resolution requires in section (4) (a) (1) that the President must submit a report to Congress within 48 hours following any case involved the United States Armed Forces.
The report must include information such as circumstances, legal authority, and the estimated scope of the involvement.
The section referenced in the 2001 AUMF resolution, Section (5) (b), is the section that requires the President to get specific authorization from Congress within a 60 day period after submitting the report.
This authorization can either come in the form of a declaration of war or a specific authorization the use the Armed Forces, and can only be waived if Congress grants an extension or is physically unable to meet \autocite{zablocki1973}.
The 2001 AUMF resolution also references the War Powers Resolution in section (2) (b) (2) by stating that ``nothing in this resolution supercedes any requirement of the War Powers Resolution.''
Senator Byrd comments on these invocations of the War Powers Resolution by saying that ``extended operations against other parties or nations not involved in the attack would require —or would it— additional specific authorization beyond the 60 day period provided for in the War Powers Resolution. Whether the language of S.J. Res. 23 adequately supports the intent is another matter \autocite[SS950]{byrd2001a}.''

>>>>

\section*{Origins of the 2001 Bush Administration AUMF}
\subsection*{The Intent of the Bush Administration}
Following the terrorist attacks against the United States that took place on September 11th, 2001, President George W. Bush role as the acting Commander in Chief and spokesman for the United States was tasked with both identifying the perpetrators of the attacks as well as deciding what counter-measures to take as a result.
President Bush identified two main groups as the parties responsible for the attacks: Al-Qaeda and the Taliban.
It was believed that Al-Qaeda was the group that planned and initiated the attacks, while the Afghani Taliban were also guilty due to harboring Al-Qaeda members within their territory.
With the purposes of providing an immediate response that would effectively utilize Congress's constitutional war powers, the 107th Congress passed an Authorization for Use of Military Force (AUMF) resolution, similar to the one passed in 1991 under the George H.W. Bush administration.
This resolution was passed through both the Senate and House in identical form on September 14th, just three days after the attacks \autocite{zotero-165}.
Although that length of time is notably fast compared to many other landmark pieces of legislation passed through Congress, there was still a lot of discussion between the legislators both of the political parties, leaders in the Bush administration, as well as legal and foreign policy experts.

@@

\noindent
By doing this, President Bush framed the attacks as more than just a tragic occurrence, they were a direct declaration of war on the free world.
In response to Al-Qaeda's act of war, the Bush administration decided to push for a legal authorization for a war against the 9/11 perpetrators.
Vice President Dick Cheney consulted with his lawyers along with members of the White House Counsel and the Department of Justice Legal Counsel to see what kind of authority would be needed.
Following this, President Bush met with congressional leaders to specifically request an resolution giving an authorization for military force to the executive branch.
Additionally, Bush also had Vice President Cheney's team of counsel members create a first draft of the resolution \autocite[176]{murray2015}.
This draft was considered by Senate Majority Leader, who would later be the sponsor for the 2001 AUMF resolution, as a blank check that would allow the Bush administration to use any amount of military force without consequence.
Congress decided to pass an AUMF under the conditions that it would limited to those responsible for the attacks, as well as that it would remain consistent with the 1973 War Powers Resolution.
The Bush administration tried to add a clause to the draft resolution that would allow for military action \emph{within} the United States, but it was rejected by Congress.
After the members of Congress agreed to a set of conditions along with the wording of the resolution, they passed the resolution through both the House and Senate in approximately 28 minutes with only a single vote in dissent \autocite[177]{murray2015}.

\subsection*{The Intent of Congress}
Although Congress went through lengthy discussions in order to make sure the wording of the resolution wouldn't have unforeseen consequences, there were still some parts that ending up being broader than intended.
The 2001 AUMF authorized the President to have access to Necessary and Proper clause in Article 1 of the Constitution in order to both combat and detain the perpetrators of the 9/11 attacks, and as such gave President Bush a large amount of military power that would typically be in the hands of Congress.
The first example is the phrase ``organizations, or persons'' which can be found in the first operative clause of the resolution.
This was dissimilar from Congress's historical declarations of war as well as the 1991 Authorization for Use of Military Force Against Iraq Resolution, which not only specifically chose the government of Iraq as a target, but also was much more clear in its goals.
Additionally, the 2001 AUMF resolution was somewhat broad in the way it described which non-state actors and/or nations should be targeted.
Malcolm Brooks Savage III describes how there was only one limitation placed on the powers of the President in the operational clauses of the 2001 AUMF resolution, being the 9/11 nexus requirement.
The first operative clause states that the President can use force against all ``nations, organizations, or persons \emph{he determines} planned, authorized, committed, or aided the terrorist attacks.''
The phrase ``he determines'' is especially notable as it could presumably allow the President to target any group that was remotely connected to the attacks, even if they are outside of the original Congressional intent of the resolution \autocite[5]{savage2015}.

The shift of military focus from state to non-state actors marked a drastic change within the foreign policy of the United States.
Although the Bush administration specifically identified Al-Qaeda and the Taliban as the targets of their response plans, the language of the 2001 AUMF resolution gave additional power to the President that went beyond those two groups specifically.
Additionally, the phrasing led to interpretations that would allow military force to be exerted not just on group actors, but also individuals.
As a result of this, President Bush as well as all future presidents that chose to implement the 2001 AUMF into their foreign policy decisions were able to choose targets in both a wider scope due to the lack of specificity and increased executive discretion, as well as choose targets that were much more specific, such as specific people that were deemed as terrorists \autocite[3]{weed2015}.
As will be discussed later in this paper, these parts of the 2001 AUMF, regardless of whether they should be considered features or loopholes, have had many direct consequences on the foreign policy of the United States for the last two decades.

@@

\begin{displayquote}
``I began to have some qualms over how broad a grant of authority Congress
gave him in our rush to act quickly.
Because of the speed with which it was passed, there was little discussion establishing a foundation for the resolution.
Because of the paucity of debate, it would be difficult to glean from the
record the specific intent of Congress in approving S.J. Res. 23.
There were after-the-fact statements made in the Senate, and there was some debate in the House, but there was not the normal level of discussion or the normal
level of analysis of the language prior to the vote that we have come to expect in the Senate \autocite[S9949]{byrd2001a}.''
\end{displayquote}

\noindent
This proves how despite the level of deliberation taken, it was still rushed far beyond the amount expected out of such an important piece of legislation.
Although there was supposedly some benefit to this, being that the situation at hand demanded immediate action, this gives evidence to the idea that the text of the resolution wasn't completely aligned with the interests or intents of Congress.

\begin{displayquote}
``Two aspects of the resolution are key: First, the use of force authority granted to the President extends only to the perpetrators of the September 11 attack.
It was not the intent of Congress to give the President unbridled authority—I hope it wasn’t to wage war against terrorism writ large without the advice and consent of Congress.
That intent was made clear when Senators modified the text of the resolution proposed by the White House to limit the grant of authority to the September 11 attack \autocite[S9949]{byrd2001a}.''
\end{displayquote}

\noindent
This quote describes the intent behind the 9/11 nexus requirement that was discussed by Savage.
Later in the speech, Senator Byrd elaborates on the resolution draft that was proposed by the White House.
This draft, in addition to all of the powers granted in the version of the resolution which was later passed by Congress, would have allowed for the President to ``deter and pre-empt any future acts of terrorism or aggression against the United States \autocite[S9949]{byrd2001a}.''
The rejection of this proposed clause proves the congressional intent to limit the scope of the authorization.

Beyond this, the resolution also aimed to further the restrictions on the power of the executive branch by invoking the War Powers Resolution, which was passed under President Nixon in 1973 \autocite{zotero-176}.
Section (2) (b) (1) of the resolution requires that specific statutory authorization should follow the meaning of section (5) (b) of the War Powers Resolution \autocite{daschle2001}.
The War Powers Resolution requires in section (4) (a) (1) that the President must submit a report to Congress within 48 hours following any case involved the United States Armed Forces.
The report must include information such as circumstances, legal authority, and the estimated scope of the involvement.
The section referenced in the 2001 AUMF resolution, section (5) (b), is the section that requires the President to get specific authorization from Congress within a 60 day period after submitting the report.
This authorization can either come in the form of a declaration of war or a specific authorization the use the Armed Forces, and can only be waived if Congress grants an extension or is physically unable to meet \autocite{zablocki1973}.
The 2001 AUMF resolution also references the War Powers Resolution in section (2) (b) (2) by stating that ``nothing in this resolution supercedes any requirement of the War Powers Resolution.''
Senator Byrd comments on these invocations of the War Powers Resolution by saying that ``extended operations against other parties or nations not involved in the attack would require —or would it— additional specific authorization beyond the 60 day period provided for in the War Powers Resolution. Whether the language of S.J. Res. 23 adequately supports the intent is another matter \autocite[SS950]{byrd2001a}.''

\section*{The Obama Era}
\subsection*{Initial Choices and Theories}
Following the end of the Bush administration and the newly forming Obama administration, the wars on terror was passed directly in President Obama's hands.
Compared to prior administrations of Reagan, Clinton, and Bush, President Obama had much less freedom in regards to choosing how to operate American foreign policy as the Commander-in-Chief.
As described by Starr-Deelen, the Bush administration's unique approach to counterterrorism left behind a \textit{fait accompli} war, meaning that the consequences of the war would largely fall under the Obama administration even though President Obama had no choice but to accept them.
During the course of his presidency, President Bush shifted the United States away from the law enforcement approach to counterterrorism in favor of the armed conflict approach \autocite[2]{starr-deelen2018}.
The law enforcement paradigm, which had historically been favored prior to the Bush administration, seeks to treat terrorism as a criminal act rather than a national security issue.
Supporters of this approach believe that terrorism is distinct from other violent acts such as war or invasion.
Counterterrorism tends to be more focused and centralized under this paradigm, as individual terrorists or specific terrorist groups are to be targeted by policymakers on a case-by-case basis.
The policy focus also tends to put more of a focus on domestic protection and preemption rather than international military force.
Additionally, the law enforcement model would place more of an emphasis on law enforcement mechanisms rather than military force.
This would optimally result in terrorists being apprehended by police forces and then would seek to implement criminal trials with prosecutors and civilian juries.
Through this it would also seek to follow principles of due process, just like any other severe criminal act \autocite[3]{starr-deelen2018}.

The main alternative to the law enforcement paradigm would be the armed conflict approach.
This paradigm treats counterterrorism as a national security threat rather than a criminal act.
Supporters believe that the existence of terrorist acts is more of an attack on the state apparatus than an attack on people.
Under this approach, terrorists should be fought and apprehended by military forces by any means, notably including lethal force.
When captured, the approach is also much more willing to waive principles of due process through the use of military courts and petitions.
As a result of this, the paradigm is also known as the ``war on terror'' approach.
Counterterrorism under this paradigm tends to follow a more international and aggressive approach where it tries to directly combat and destroy terrorism regardless of geography or boundaries.
In regards to framing, the approach tends to contrast from the law enforcement approach by labeling terrorists as enemy combatants rather than criminals.
The Bush administration was a strong proponent of the armed conflict approach, as it labeled jihadists in Al-Qaeda and the Taliban as soldiers in an ideological and violent war against the United States and the broader West \autocite[3]{starr-deelen2018}.

In response to the war on terror approach, the newly formed Obama administration would have three options.
First, President Obama could embrace the armed conflict paradigm, but announce the end of the war on terror.
This approach would attempt to simply move on from the Bush Era and create a clean slate.
Second, President Obama could take a more aggressive approach against the Bush administration's approach by renouncing the war on terror and taking action against the legal violations of the Bush administration.
This approach would seek to likely seek to return back to the prior status quo of law enforcement based counterterrorism.
Finally, President Obama could attempt to create a third path by combining aspects of both the law enforcement and armed conflict paradigms.
President Obama would end up deciding to take the third approach that would keep the fight against the terrorism while reducing executive overreach and returning to a constitutional balance of power.
Upon taking office, President Obama stated that the President does not have the power to ``unilaterally authorize a military attack in a situation that does not involve stopping an actual or imminent threat to the nation \autocite[6]{starr-deelen2018}.''
He also signed executive orders to close the Guantanamo Bay prison, end the use of torture in interrogations, and review prior detention policies.
President Obama's deputy nation security adviser also echoed this by planning to implement a model that would incorporate both the criminal justice system as well as international military force.
This new approach would also seek to shift the focus toward combating the conditions behind extremism rather than the extremist terrorists themselves \autocite[7]{starr-deelen2018}.

\subsection*{Expansions of the AUMF Scope}
Although this third way policy was implemented to a certain degree, the Obama administration's use of the 2001 AUMF resolution involved stretching its application far beyond the original Congressional intent.
Specifically, the authorization started to be applied to groups and conflicts far beyond those directly involved in the 9/11 terrorist attacks.
The primary way that this was done was by expanded the target of executive force to groups that could be associated with the Al-Qaeda in Afghanistan.
By designating groups as \emph{co-belligerents} with the Afghani Al-Qaeda, the Obama administration was effectively able to designate groups that weren't involved in the 9/11 attacks as being covered by the 9/11 nexus.
This could even include groups that had not been formed until after 2001.
During President Obama's first term, the administration first targeted Al-Qaeda in the Arabian Peninsula (AQAP), which was primarily located in Yemen.
Additionally, other non-Al-Qaeda groups such as al-Shabaab in Somalia and the Khorasan Group in Syria were also likely considered to be targets of the 2001 AUMF \autocite[189]{murray2015}.

Part of this expansion in executive power was an expansion of ambiguity and secrecy around Obama administration actions.
The drone strike program under the Obama administration wasn't officially acknowledged for three years despite estimates that more strikes took place in Pakistan during President Obama's first year in office than the amount under Bush's previous seven years in office \autocite[189]{murray2015}.
This included a secret legal decision in the Obama administration's Office of Legal Counsel in 2010 which allowed for the targeted killing of American citizens abroad as cobelligerents. 
This resulted in the drone killing of Anwar Awlaki, a US citizen that was associated with AQAP \autocite[191]{murray2015}.
The marked a shift back to the armed conflict approach of the Bush administration, as it became clear that lethal force and a lack of due process could be used at the discretion of the executive branch.

To achieve these expansion, the Obama administration regularly cited the 2001 AUMF as a legal basis.
They have further claimed that the 2001 AUMF resolution has no limits in regards to geography or temporal scale \autocite[108]{boyle2015}.
This claim, if followed, would lead to a scope of power that goes far beyond that of the original authors of the 2001 AUMF resolution.
In regards to the time scale, it would be reasonable to assume that this interpretation would leave to the power of the authorization in the hands of the president until Al-Qaeda is completely defeated.
Due to their expansion around the world, Obama administration officials have estimated that the war on terror could end up lasting for upwards of 30 years.
Additionally, it could also extend beyond Al-Qaeda, as it was commonly used to justify military action against ISIS in Iraq and Syria despite the fact that ISIS did not exist during the 9/11 attacks and also has a ``remarkably loose affiliation'' with Al-Qaeda \autocite[111]{boyle2015}
The interpretation would also give legal legitimacy to targeted attacks in any geographic location where links could be made to Al-Qaeda or the Taliban.
There is reason to believe that this was used to justify Obama administration attacks not just in other Middle Eastern nations such as Iraq and Libya, but also in other regions in Asia such as Indonesia and the Philippines.
It is unknown exactly where the Obama administration has invoked the 2001 AUMF as many attacks were either unrecognized or done under private criteria set by the administration \autocite[109]{boyle2015}.

\subsection*{The Preservation of Executive Detention Policy}
Although President Obama remained in favor of closing the Guantanamo Bay detention center throughout his time in office, the actions of the Obama administration legal team suggest that he was still in favor of continuing to detain prisoners at the site without civil trial, albeit with some different requirements from those under President Bush.
Most of this position can be explained with a memorandum known as the ``Obama Memo'' in which parts of the Bush administration's position on the detention of suspected and known terrorists were both preserved, refuted, and adjusted.
The main way that the memo differed from the Bush administration's stance was by removing any mention to Article II of the Constitution, which was previously interpreted as giving the President an inherent right to unilateral detention power.
This would presumably give some of the power of detention back to Congress, and would allow for Congress to clarify whether or not an authorization for force would include detention authority \autocite[44]{brill2010}.
The Obama Memo also seemed to invoke a further care for international law, as it contradicted the Bush administration's arguments that international treaties such as the Geneva Conventions were limited and didn't apply to their actions.
Specifically, the memo referenced the United Nations Charter, NATO, the OAS, the Geneva Conventions, and the International Committee of the Red Cross \autocite[46]{brill2010}.
The result of these changes included the limiting the power of the executive branch by providing oversight from both the legislative branch as well as the international community.
These changes were presumably meant to aid in the eventual closing of Guantanamo detention facilities as they would restrict the ability to bring new groups of detainees into the scope of executive detention power.

Despite these proposals, the Obama administration remained remarkably similar in policy to the Bush administration when it came to the detainees that were already present in the facilities during President Obama's two terms in office.
First, the Obama Memo still used an interpretation of the 2001 AUMF as a legal justification for detaining the prisoners.
In the section regarding who could be detained by the executive branch, the memo states that the President can detain people who ``were part of, or substantially supported, Taliban or al Qaeda forces or associated forces.''
This was virtually identical to the specifications of the Bush administration apart from the word ``substantially''.
The inclusion of that word likely provided no changes to the policy of detention as there was no definitions or clarifications for what that word would mean \autocite[46]{brill2010}.

Additionally, there is reason to believe that the changes to include international law and decision-makers in the detention process were more aesthetic than substantial.
While the administration spoke to the relevance of international law in the interpretation and application of the 2001 AUMF, the actual interpretations of the law generally increased presidential discretion.
In some cases, the Obama administration even attempted to invoke international law as a reason to increase executive power.

\begin{displayquote}
``On many of the most controversial issues of presidential power under the AUMF — including the legality of indefinite military detention of persons captured away from a traditional battlefield, the targeting and capturing of alleged terrorists who are not formal members of an identified enemy group, and the use of force in nations with which the United States is not at war — the administration justified its actions by reference to international law \autocite[639]{bradley2016}.''
\end{displayquote}

\noindent
In addition, the Obama administration argued that the 2001 AUMF granted the president the power to detain suspected terrorists even if they tried to flee battle or conceal themselves as a civilian.
This directly contradicts the International Committee of the Red Cross which believing that detention authority should be limited to ``those who directly participate in hostilities \autocite[639]{bradley2016}'', even though the committee was cited in the Obama Memo.
An additional sign of executive discretion could be found in the aforementioned expansion scope of who could be detained, since it would allow for the president to detain anyone who they deemed as being a supporter of Al Qaeda through the principle of co-belligerency, despite there being no precedent for this type of proceeding in a non-international conflict against decentralized targets.
Although the brief stated that principles of international law would have to inform the interpretations of the 2001 AUMF, it provided no further clarifications on what principles it was referring to or how they would have to inform the legal decisions.

Despite these factors however, some would still argue that the Obama administration's interpretations of detention law still have some substantial changes from the Bush administration, even if the changes weren't immediate.
Brill states that the international law arguments were structured in a way that would allow for future changes to the presumably unlawful executive detention system without having to undo the order that was previously created and held throughout the Bush presidency and the early parts of the Obama presidency.

\begin{displayquote}
``The AUMF and international law arguments, in other words, are quintessentially minimalist. They preserve a status quo with regard to particular detainees but contain no broader ambitions for enlarging executive power, asserting an isolationist approach to international law, or reserving the right to ignore Congress. And they are quintessentially realist in they see the policy context of an ongoing effort to clean up a difficult mess as germane to their legal claims about what the AUMF means. In short, the Obama Memo shows an administration trying to maintain some sort of order as it unwinds an institution whose closure was written into an Executive Order on the President’s first full day in office \autocite[48]{brill2010}.''
\end{displayquote}

\noindent
This argument would explain the apparent gap between President Obama's outward policy promise to close Guantanamo Bay and end the unjust executive detentions while internally creating and supporting legal interpretations that allowed the detention center to stay open.

\section*{The Trump Era}
Although this section will be more brief due to there being less research and documentation on the Trump Administration as well as the fact that there weren't significant changes from the Obama Administration in terms of legal interpretations, it is still worth noting the relevance of the 2001 AUMF to the Trump administration's foreign policy.
The Trump campaign in 2016 focused much less on foreign military policy compared to both of the Obama campaigns, but President Trump still seemed to share some of the same rhetoric of getting us out of the ``endless wars'' and bringing the troops home as part of his America First proposal.
Simultaneously, President Trump also followed a rhetoric that was highly critical of Obama's national security decisions regarding the threats of ISIS, Al-Qaeda, and ``radical Islamic terrorism'' as a whole.
Despite this, he didn't have many specific policy prescriptions in mind compared to Obama's proposals such as closing down Guantanamo Bay and passing a new authorization of military force.
Additionally, the lack of political experience led to there being more of an emphasis among political scholars on the foreign policy staff of the Trump Administration rather than Donald Trump himself \autocite[24]{starr-deelen2018}.

It seems as if the changes in foreign policy have been more along the lines of application of prior legal interpretations rather than substantial changes in the interpretations themselves.
President Trump has also continued the use of drone strikes outside of the original 9/11 theaters, as was seen in the Obama presidency \autocite[35]{sterio2018}.
Along these lines, the Trump Administration has legally agreed with the prior Bush and Obama administration interpretations of the 2001 AUMF in regards to drone strikes.
The willingness to use lethal force on suspected terrorists, as shown by both the drone strikes as well as specific incidents such as the assassination of Iran's General Qassem Soleimani, demonstrate that Trump has been more than willing to stick to the Bush Administrations armed conflict paradigm instead of attempting to shift approaches like the Obama Administration.
The Trump administration the covert and secretive nature of force application that was found in both of the prior administrations.
